\large{Vipassana for Hackers} \\ \Huge{Paper Two: The Brain} \\ \large\textit{Version 0.2}

### 2019-10-10 paper-two/paper-two.tex
edits to preethis most recent fixes

Perhaps the strangest experience during a Vipassana course is that of shifts in the
sense of self. The ``self'', as a subjective agent in space and time, changes. At
first, this change is only very slight. As a student closes her eyes and begins to
meditate, it may feel as though the body has shifted in space or rotated in different direction, this is
due to the temporary down regulation of the prefrontal cortex. As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the AIC
fires more rapidly, leading to alternative networks of consciousness
arising. \todo

[Side note; Consciousness is defined as our subjective awareness of ourselves and our environment. 
An altered state of consciousness is defined as alternate patterns or configurations of 
this experience, which differ qualitatively from a baseline state.
Altered states of consciousness include a state of dreaming, epileptic seizures, sensory deprivation, 
hysterical states of dissociation and depersonalization as well as pharmacologically induced altered states.

The entropic brain hypothesis: This was put forth by Robin Carhatt Harris et al, while studying 
the psychedelic state. Entropy is a measure of the uncertainty in a system and in this case applied 
to the states of consciousness and their associated neurodynamics. The theory makes a distinction 
two fundamental modes of consciousness: Primitive consciousness and the normal waking or secondary
consciousness. 
Primitive consciousness is characterized by higher entropy and it is hypothesized to have 
preceded the development of modern, adult, human, normal waking consciousness. 
It is argued that suppression of this entropy facilitates normal waking consciousness 
and associated metacognitive functions such as reality-testing and self-awareness.]

Sometimes it feels like loosening the hold of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

\bibitem{experiencemeda}
  Zoran Josipovic, Ilan Dinstein, Jochen Weber, David J. Heeger
  \textit {Influence of meditation on anti-correlated networks in the brain}
  Front. Hum. Neurosci., 03 January 2012,
  https://doi.org/10.3389/fnhum.2011.00183

>>>>

Perhaps the strangest experience during a Vipassana course is that of shifts in the
sense of self. The ``self'', as a subjective agent in space and time, changes. At
first, this change is only very slight. As a student closes her eyes and begins to
meditate, it may feel as though the body has shifted in space or rotated in different direction, this is
due to the temporary down regulation of the prefrontal cortex. As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the AIC
fires more rapidly, leading to alternative networks of consciousness
arising.

Consciousness is defined as our subjective awareness of ourselves and our environment.
An altered state of consciousness is defined as alternate patterns or configurations of
this experience, which differ qualitatively from a baseline state.
Altered states of consciousness include a state of dreaming, epileptic seizures, sensory deprivation,
hysterical states of dissociation and depersonalization as well as pharmacologically induced altered states.

The notion of altered states of consciousness has been put forth another way by the
Entropic Brain Hypothesis, described by Robin Carhart-Harris et al, while studying
the brain under the influence of psychedelics. \cite{entropic,entropicrevisited} Entropy is a measure of the
uncertainty in a system and, in this case, applies
to states of consciousness and their associated neurodynamics. The theory makes a
distinction between
two fundamental modes of consciousness: Primitive Consciousness and the normal Secondary
Consciousness of waking states.
Primitive consciousness is characterized by higher entropy and it is hypothesized to have
preceded the development of modern, adult, human, normal waking consciousness.
It is argued that suppression of this entropy facilitates normal waking consciousness
and associated metacognitive functions such as reality-testing and self-awareness.

Even one ten-day course will expose a student to these altered states of
consciousness and provide access to Primitive Consciousness. However, unlike
psychedelics, the Vipassana student is not under the control of an external force. If
she opens her eyes and stops meditating, these disorienting altered states of
consciousness instantly disintegrate. Lasting changes to consciousness caused by
Vipassana take a different shape.

Sometimes it feels like loosening the hold of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to the influence of certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of
ambiguity. \cite{cortexbliss} Outside the activity of meditation, the lasting effects
(``Altered Traits'') of meditation are not so extreme. They tend to include
heightened awareness, improved concentration, and increased empathy, among
others. \cite{alteredtraits}

@@

\bibitem{experiencemeda}
  Zoran Josipovic, Ilan Dinstein, Jochen Weber, David J. Heeger
  \textit {Influence of meditation on anti-correlated networks in the brain}
  Frontiers in Human Neuroscience, 03 January 2012,
  https://doi.org/10.3389/fnhum.2011.00183

\bibitem{entropic}
  Robin L. Carhart-Harris, Robert Leech, Peter J. Hellyer, Murray Shanahan, Amanda
  Feilding, Enzo Tagliazucchi, Dante R. Chialvo and David Nutt
  \textit{The entropic brain: a theory of conscious states informed by neuroimaging
    research with psychedelic drugs}
  Frontiers in Human Neuroscience, 03 February 2014,
  https://doi.org/10.3389/fnhum.2014.00020

\bibitem{entropicrevisited}
  Carhart-Harris, Robin L.
  \textit{The entropic brain - revisited}
  NEUROPHARMACOLOGY, Vol: 142, Pages: 167-178, ISSN: 0028-3908, 2018.
